# `sf` package: quick references

The `sf ` package provides a set of tools for working with geospatial data in R. 
For details and a look under the hood of this package, see: **[sf github page](https://github.com/r-spatial/sf)**
 
Huge shout out to the creators of the `sf` package and those who continue to improve upon it - for making spatial analyses free and accessible and feasible: 


:::{.citationbox}
::::{.citationbox-header}
:::::{.citationbox-icon}
:::::
🙏 `sf` creators:
::::
::::{.citationbox-body}
  * Pebesma, E., & Bivand, R. (2023). Spatial Data Science: With Applications in R. Chapman and Hall/CRC.
  https://doi.org/10.1201/9780429459016

  * Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data. The R Journal 10 (1),
  439-446, https://doi.org/10.32614/RJ-2018-009
::::
:::


---------

Below is a subset of the most common `sf` functions that I use for basic mapping with simple features. For a complete list, check out this reference guide to download and keep: **[sf package cheat sheet](https://rstudio.github.io/cheatsheets/sf.pdf)** 

Also, this: **[online `sf` guide:](https://r-spatial.github.io/sf/index.html)**

You can run this code to get an output of all possible methods to pass an `sf` object to: 

```{r, eval = FALSE}

methods(class = "sf")

```

\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 

##### `st_read` {.unnumbered}
Use this to bring in a saved shapefile within a directory on your drive


```{r, results = "hide", message = FALSE, warning = FALSE}

library(sf) 

MT_counties <- st_read("Shapefiles/MT_counties.shp", "MT_counties")

```

\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 

##### `st_as_sf()` {.unnumbered}
Creates an `sf` object from a non-geospatial dataframe 


\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_crs()` {.unnumbered}
Retrieve coordinate reference system from sf or sfc object

* add `$proj4string` for a more concise report


```{r}

st_crs(MT_counties)$proj4string

```

\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_transform()` {.unnumbered}
Transform or convert coordinates of simple feature; set the coordinate reference system (crs) using EPSG code

```{r}

MT_counties_projected <- st_transform(MT_counties, crs = 32100)

st_crs(MT_counties_projected)$proj4string

```

**UTM Zone Projections:**

```{r}
MT_counties_projected_zone11 <- st_transform(MT_counties, crs = 26911)
MT_counties_projected_zone12 <- st_transform(MT_counties, crs = 26912)

```



```{r}

library(ggplot2)
library(cowplot)

unprojected <- ggplot() + 
  geom_sf(data = MT_counties, fill = "grey90") + 
  ggtitle("unprojected (GCS)") +
  theme_void()

projected <- ggplot() + 
  geom_sf(data = MT_counties_projected, fill = "grey90") + 
  ggtitle("projected (ESPG:32100)") +
  theme_void()

projected_zone11 <- ggplot() + 
  geom_sf(data = MT_counties_projected_zone11, fill = "grey90") +
  ggtitle("projected for Zone 11N (ESPG:26911)") +
  theme_void()

projected_zone12 <- ggplot() + 
  geom_sf(data = MT_counties_projected_zone12, fill = "grey90") +
  ggtitle("projected for Zone 12N (ESPG:26912)") +
  theme_void()

# Create the 2x2 grid of plots
grid_plot <- plot_grid(
  unprojected, projected, 
  projected_zone11, projected_zone12, 
  ncol = 2, nrow = 2, align = "hv"
)

# Add vertical and horizontal separators
grid_plot +
  draw_line(x = c(0.5, 0.5), y = c(0, 1), color = "black", size = 0.5) + # Vertical line
  draw_line(x = c(0, 1), y = c(0.5, 0.5), color = "black", size = 0.5)   # Horizontal line


```






\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 

##### `st_centroid()` {.unnumbered}

```{r}




```



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_bbox()` {.unnumbered}



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 





##### `st_make_grid()` {.unnumbered}



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_buffer()` {.unnumbered}



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_crop()` {.unnumbered}



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 

---------- 


##### `st_join()` {.unnumbered}

Performs a spatial left or inner join between "x" and "y"



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_union()` {.unnumbered}
creates a single geometry from multiple geometries, consisting of all geometry elements

\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 

##### `st_intersection()` {.unnumbered}
creates a geometry from the shared portion of "x" and "y"



\newpage 

```{r, results='asis', echo = FALSE}

cat('<hr style="border: none; margin-bottom: 30px;">')

```
 
---------- 


##### `st_difference()` {.unnumbered}
creates a gemoetry from "x" that **does not** intersect with "y"







