# Rasters

```{r, echo = FALSE}

source("_common.R")

```


To recap from previous chapters: 
**Vector data** is made up of points, lines, and polygons
**Raster data** is made up of points or grid cells with assigned values

## Raster Formats {.unnumbered}

```{r, echo = FALSE}

raster_formats <- read.csv(file = here::here("info_tables/raster_formats.csv"))

raster_formats_table <- raster_formats %>%
  kbl(caption = "Raster Formats",
      align = "llccl") %>%
  column_spec(1, bold = TRUE) %>%  
  row_spec(0, font_size = 20, bold = TRUE) %>%
  kable_styling(
    bootstrap_options = c("striped"),
    full_width = TRUE, 
    position = "center", 
    font_size = 18)

raster_formats_table

```

Rasters can be stored and downloaded in a variety of formats, most commonly: 

* **.TIFF:** Tagged Image File Format
  + one of the most widely used in GIS and remote sensing
  + supports high-resolution images and can store **multiple bands** (RGB, infrared, DEMs, etc.)
  + often **georeferenced** (GeoTIFF) for spatial analysis in GIS software
  + use cases: satellite imagery and aerial photos, Digital Elevation Models (DEM), land cover classification and environmental modellin
  + Cons: can result in very large file sizes
  
* **.netCDF:** Network Common Data Form 
  + a self-describing format used for multidimensional data
  + commonly used in climate modeling, atmospheric research, and oceanographic data
  + supports **multiple dimensions** (latitude, longitude, altitude, time)
  + handles large multidimensional dataets (time series, 3D spatial models)
  + more efficient storage and compression (HDF-based structure)
  + highly compatible with computing tools and software (R, Python, MATLAB, C, etc.)
  + Cons: not always natively supported by GIS software (might need to convert to a GeoTIFF); steeper learning curve for beginners 
  + use cases: climate models and weather forecasting (NASA and NOAA datasets, e.g.); sea surface temperature, wind patterns, other oceanic data; long-term environmental monitoring (e.g. precipitation trends over time); 3D spatial models (e.g. atmospheric data with altitude layers integrated) 
  
* **.JPG:** Joint Photographic Experts Group
  + lossy-compresssed raster format that reduces file size but sacrifices image quality
  + highly compressed with small file size
  + good for web mapping applications
  + widely supported across platforms
  + does not support multi-band or georeferencing
  + unsuitable for spatial analyses but good for display (such as using as a basemap)
  
* **.PNG:** Portable Network Graphics
  + lossless raster format
  + often used for web visualization and cartographic outputs
  + supports transparency (useful for overlays in maps)
  + larger file size compared to .JPG
  + does not support multi-band or georeferencing
  + unsuitable for spatial analyses but good for display (such as using as a basemap)



---------------

**The challenge:** Figuring out when raster data is useful or not to your mapping goals 

The same real- world scenario can be represented in both a raster or a vector representation.

Examples of information that is best represented as a raster: 

* **Elevation**
* **Terrain**
* **Land cover**
* **Disease risk or incidence?** 

  + typically, we are only ever able to visualize disease incidence and risk as vectorized data, generalized to arbitrary and artificial administrative boundaries. However, it we have the individual-level point data, rasterizing disease risk can provide a very interesting visualization as it will transcend issues related to the **MAUP** (link to bias chapter here) bias problems. 
  
  + rasterizing diseases can enable us to see the fine scale variation of risk across a smooth surface, and
  
  + we can generate rasters of disease risk and assess them as potential functions of other variables, such as investigating the raster of malaria cases as a function of temperature. In this case, it *would* make sense to keep this data as a raster. 